Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-08 manuscript/manuscript.tex
changes

Second, we reanalyzed the Longshanks selection experiment, which selected for
longer tibiae length relative to body size in mice, leading to a response to
selection of about 5 standard deviations over the course of twenty generations
\parencite{Marchini2014-de,Castro2019-uk}. This study includes two independent
selection lines, Longshanks 1 and 2 (LS1 and LS2), and an unselected control
line (Ctrl).  Consequently, this selection experiment offers a useful control
to test our between-replicate covariances: we expect to see positive
between-replicate covariance in the comparison between the two Longshanks
selection lines, but not between the two pairwise comparisons between the
control line and the two Longshanks lines. We find that this is the case (gray
confidence intervals in Figure \ref{fig:figure-2} C), with the two Longshanks
comparisons to the control line not being significantly different from zero,
while the comparison between the two Longshanks lines is statistically
significantly different from zero (CIs [0.013, 0.037]).

\vb{The presence of an unselected control line provides an alternative way to
  partition the effects of linked selection and genetic drift: we can compare
  the total variance in allele frequency change of the control line (which
  excludes the effect of artificial selection on allele frequencies) to the
  total variance in frequency change of the Longshanks selection lines. We can
  partition the allele frequency change between the two timepoints (20
  generations) for a Longshanks line as $\Delta p_{t,\mathrm{LS1}} =
  \Delta_{_D} p_{t,\mathrm{LS1}} + \Delta_{_{U}} p_{t,\mathrm{LS1}} +
  \Delta_{_\mathrm{LS}} p_t$ (and similarly for the Longshanks two line, LS2),
  and $\Delta p_{t,\mathrm{Ctrl}} = \Delta_{_D} p_{t,\mathrm{Ctrl}} +
  \Delta_{_{U}} p_{t,\mathrm{Ctrl}} + \Delta_{_\mathrm{Ctrl}} p_t$ for the
  control line. Here, $\Delta_{_D} p_{t}$, $\Delta_{_U} p_t$,
  $\Delta_{_\mathrm{LS}} p_t$ are the allele frequency changes due to drift,
  selection unique to a replicate, and selection response shared between the
  two Longshanks replicates, respectively. By construction each of these terms
  are uncorrelated, allowing us to estimate the increase in variance in allele
  frequency change due to selection as $(\var(\Delta p_{t,\mathrm{LS1}}) +
  \var(\Delta p_{t,\mathrm{LS2}}))/2 - \var(\Delta p_{t,\mathrm{Ctrl}})$ and
  the shared effect of selection across selected lines as $\cov(\Delta
  p_{t,\mathrm{LS1}}, \Delta p_{t,\mathrm{LS2}})$. We estimate at least 32\%
    (95\% CI $[21\%, 48\%]$) of the variance in allele frequency change is
    driven by the effects of selection, of which 14\% (95\% CI $[3\%, 33\%]$)
    is estimated to be unique to a selection line, and 17\% (95\% CI $[9\%,
  23\%]$) is the effect of shared selection between the two Longshanks
selection lines.}

A major finding in the Longshanks study were two major-effect loci that showed
parallel frequency shifts between the two selection lines: a region harboring
the gene Nkx3-2 known to be involved in limb development, and another region
harboring six other candidate genes. We were curious to what extent our
genome-wide covariances were being driven by these two outlier large-effect
loci, so we excluded them from the analysis. Since we do not know the extent to
which linkage disequilibrium around these large-effect loci affects neighboring
loci, we took the conservative precaution of excluding the entire chromosomes
these loci reside on (chromosomes 5 and 10), and re-calculating the temporal
covariances. We find excluding these large effect loci has little impact on the
confidence intervals (blue confidence intervals in Figure \ref{fig:figure-2}),
indicating that these across-replicate covariances are indeed driven by a large
number of loci. \vb{This is consistent with a signal of selection on a
  polygenic trait driving genome-wide change, although we note that
  large-effect loci can contribute to the indirect change at unlinked loci
\parencite{Robertson1961-ho,Santiago1995-hx}. }

>>>>

\gc{Next,} we reanalyzed the Longshanks selection experiment, which selected for
longer tibiae length relative to body size in mice, leading to a response to
selection of about 5 standard deviations over the course of twenty generations
\parencite{Marchini2014-de,Castro2019-uk}. This study includes two independent
selection lines, Longshanks 1 and 2 (LS1 and LS2), and an unselected control
line (Ctrl).  Consequently, this selection experiment offers a useful control
to test our between-replicate covariances: we expect to see positive
between-replicate covariance in the comparison between the two Longshanks
selection lines, but not between the two pairwise comparisons between the
control line and the two Longshanks lines. We find that this is the case (gray
confidence intervals in Figure \ref{fig:figure-2} C), with the two Longshanks
comparisons to the control line not being significantly different from zero,
while the comparison between the two Longshanks lines is statistically
significantly different from zero (CIs [0.013, 0.037]).

\gc{One} finding in the Longshanks study were two major-effect loci that showed
parallel frequency shifts between the two selection lines: a region harboring
the gene Nkx3-2 known to be involved in limb development, and another region
harboring six other candidate genes. We were curious to what extent our
genome-wide covariances were being driven by these two outlier large-effect
loci, so we excluded them from the analysis. Since we do not know the extent to
which linkage disequilibrium around these large-effect loci affects neighboring
loci, we took the conservative precaution of excluding the entire chromosomes
these loci reside on (chromosomes 5 and 10), and re-calculating the temporal
covariances. We find excluding these large effect loci has little impact on the
confidence intervals (blue confidence intervals in Figure \ref{fig:figure-2}),
indicating that these across-replicate covariances are indeed driven by a large
number of loci. This is consistent with a signal of selection on a
  polygenic trait driving genome-wide change, although we note that
  large-effect loci can contribute to the indirect change at unlinked loci
\parencite{Robertson1961-ho,Santiago1995-hx}.

\gc{The presence of an unselected control line provides an alternative way to
  partition the effects of linked selection and genetic drift: we can compare
  the total variance in allele frequency change of the control line (which
  excludes the effect of artificial selection on allele frequencies) to the
  total variance in frequency change of the Longshanks selection
  lines.
We can partition the allele frequency change between the two timepoints (20
  generations) for a Longshanks line as $\Delta p_{t,\mathrm{LS1}} =
  \Delta_{_D} p_{t,\mathrm{LS1}} + \Delta_{_{U1}} p_{t} +
  \Delta_{_\mathrm{S}} p_t$ 
  $\Delta_{_D1} p_{t}$, $\Delta_{_U} p_t$,
  $\Delta_{_\mathrm{LS}} p_t$  the allele frequency change due to
  drift in replicate 1 ($D1$),
  selection unique to the 1$^{st}$ LS replicate ($U1$), and selection response shared between the
  two Longshanks replicates (LS) respectively (and similarly for the Longshanks two line, LS2).
  %partition the allele frequency change between the two timepoints (20
  %generations) for a Longshanks line as $\Delta p_{t,\mathrm{LS1}} =
  % \Delta_{_D} p_{t,\mathrm{LS1}} + \Delta_{_{U}} p_{t,\mathrm{LS1}} +
  % \Delta_{_\mathrm{LS}} p_t$ (and similarly for the Longshanks two line, LS2),
  % and $\Delta p_{t,\mathrm{Ctrl}} = \Delta_{_D} p_{t,\mathrm{Ctrl}} +
  % \Delta_{_{U}} p_{t,\mathrm{Ctrl}} + \Delta_{_\mathrm{Ctrl}} p_t$ for the
  % control line. Here, $\Delta_{_D} p_{t}$, $\Delta_{_U} p_t$,
  % $\Delta_{_\mathrm{LS}} p_t$ are the allele frequency changes due to drift,
  % selection unique to a replicate, and selection response shared between the
  % two Longshanks replicates, respectively.
  By construction we will assume that each of these terms
  are uncorrelated, and we will assume that we can approximate the
  contribution of genetic drift in the LS
  lines the variance in allele frequency change observed in
  the control ($\var(\Delta p_{t,\mathrm{Ctrl}})
  = \var(\Delta_{_D1} p_{t})=\var(\Delta_{_D2} p_{t})$)
  This allows us to estimate the increase in variance in allele
  frequency change due to selection as $(\var(\Delta p_{t,\mathrm{LS1}}) +
  \var(\Delta p_{t,\mathrm{LS2}}))/2 - \var(\Delta p_{t,\mathrm{Ctrl}})$ and
  the shared effect of selection across selected lines as $\cov(\Delta
  p_{t,\mathrm{LS1}}, \Delta p_{t,\mathrm{LS2}})$. We estimate at least 32\%
    (95\% CI $[21\%, 48\%]$) of the variance in allele frequency change is
    driven by the effects of selection, of which 14\% (95\% CI $[3\%, 33\%]$)
    is estimated to be unique to a selection line, and 17\% (95\% CI $[9\%,
  23\%]$) is the effect of shared selection between the two Longshanks
selection lines.}
